---
title: "Big Data Computer Assessment"
output:
  html_document: default
  pdf_document: default
---

Candidate Number: 1073601

```{r, echo=FALSE}
# List of required packages
required_packages <- c(
  "mosaic", "phytools", "popbio", "popdemo", "Rage", "Rcompadre",
  "remotes", "rotl", "taxize", "tidyverse")

# Installs the missing packages
missing_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(missing_packages)) {
  install.packages(missing_packages)
}
#Loads the missing packages
lapply(required_packages, library, character.only = TRUE)
```

```{r}
#Downloading the Comadre dataset
#Fetch the most recent version
comadre <- cdb_fetch("comadre")

#Selecting Reptiles
#located worldwide and not from captivity:
comadre_rep <- subset(comadre,
                   # Consider only reptiles
                   Class == "Reptilia" & 
                   # anywhere in the world so no lat requirement
                   # studied under "control" (i.e. no treatment) conditions
                   MatrixTreatment == "Unmanipulated" &
                   # and only from "W"ild populations
                   MatrixCaptivity == "W") 
#Flag and remove matrices with missing values or dont meet assumptions of ergodicity
comadre_rep_flag <- cdb_flag(comadre_rep)
comadre_reptiles <- subset(comadre_rep_flag,
                      check_NA_A == FALSE &
                      check_ergodic == TRUE)
```

**Question 1**

Show a histogram of the population growth rate, and a separate histogram of the generation time of your subset of species. Describe the patterns. Do they make sense to you?

```{r warning=FALSE}
#Calculating the population growth rate by finding the dominant eigenvalue of each matrix, then storing this in the reptiles dataframe
comadre_reptiles$lambda <- unlist(lapply(matA(comadre_reptiles), popbio::lambda))

#Calculating gen time using the rage function
comadre_reptiles$gen_T <- mapply(Rage::gen_time, matU(comadre_reptiles), matF(comadre_reptiles))

#ensuring a clean version of the dataset for phylogenies
comadre_reptiles_clean <- comadre_reptiles |>
  filter(!is.na(lambda), 
         !is.na(gen_T), 
         !is.infinite(gen_T))
```

```{r warning = F}
#Plotting the pop growth rate histogram - red abline represents a stable population that is neither growing nor shrinking
hist(comadre_reptiles_clean$lambda, xlab= "Population growth rate", main = NULL) 
  abline(v = 1, col = 'red', lty = 2, lwd = 2)
```

```{r warning = F}
#Plotting generation time histogram
hist(comadre_reptiles_clean$gen_T, xlab = "Generation time (years)", main = NULL)
```

The pop growth rate histogram shows that the majority of populations have a $\lambda$ close to 1, suggesting relatively stable population growth rates. There are, however, a few populations that have $\lambda$ values significantly lower than 1, suggesting a declining population, perhaps due to climate change. Conversely, there is a wide variety in populations with $\lambda$ values greater than one, with some populations growing steadily but others with explosive growth, such as the population with a $\lambda$ close to 2.5. As a result, the overall distribution is asymmetrical.

Similarly, most of the generation times are below 20 years, suggesting that most studied reptiles have short reproductive time-spans. The distribution is skewed, however, with some outlier populations that have longer (20-40 years) or significantly longer (50-80 years) generation times, but the frequency of such populations declines as the generation time increases.

**Question 2**

Fit a model to explore whether the generation time and population growth rate of your species’ populations are correlated. Would you have expected them to be correlated? Why?

```{r}

# Scatter plot
ggplot(comadre_reptiles_clean, aes(x = lambda, y = gen_T )) +
  geom_point(alpha = 0.7) +  
  labs(y = "Generation Time (gen_T)", 
       x = "Population Growth Rate (lambda)", 
       title = "Population Growth Rate vs. Scatter Plot of Generation Time") +
  theme_classic()

```

From the previous question and the above exploratory scatter plot, we can see that both $\lambda$ and generation time are **not** normally distributed. Therefore, we cannot use a linear model as this would violate the assumption of normality. Instead, we will use a spearmans correlation test:

```{r warning = F}
# Spearman's correlation test
spearman_result <- cor.test(comadre_reptiles_clean$lambda, 
                            comadre_reptiles_clean$gen_T, 
                            method = "spearman")

print(spearman_result)
```
Interestingly, we observe that there is no significant correlation as p > 0.05, with a very weak negative correlation coefficient (rho) of -0.119. This is surprising as one might have expected species with higher population growth rates to have shorter generation times, which would hypothetically lead to a strong negative correlation. Given the lack of significance of the calculated correlation, however, this does not seem to be the case. 

**Question 3**

Using the IUCN API token provided (or via your own), or via the IUCN csv file provided
explore, using statistical models, whether generation time, on the one hand, and population
growth rate, on the other hand, predict the Red List conservation status of your species
subset.