---
title: "Big Data Computer Assessment"
output:
  html_document: default
  pdf_document: default
---

Candidate Number: 1073601

```{r, echo=FALSE}
# List of required packages
required_packages <- c(
  "mosaic", "phytools", "popbio", "popdemo", "Rage", "Rcompadre",
  "remotes", "rotl", "taxize", "tidyverse")

# Installs the missing packages
missing_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(missing_packages)) {
  install.packages(missing_packages)
}
#Loads the missing packages
lapply(required_packages, library, character.only = TRUE)
```

```{r}
#Downloading the Comadre dataset
#Fetch the most recent version
comadre <- cdb_fetch("comadre")

#Selecting Reptiles
#located worldwide and not from captivity:
comadre_rep <- subset(comadre,
                   # Consider only reptiles
                   Class == "Reptilia" & 
                   # anywhere in the world so no lat requirement
                   # studied under "control" (i.e. no treatment) conditions
                   MatrixTreatment == "Unmanipulated" &
                   # and only from "W"ild populations
                   MatrixCaptivity == "W") 
#Flag and remove matrices with missing values or dont meet assumptions of ergodicity
comadre_rep_flag <- cdb_flag(comadre_rep)
comadre_reptiles <- subset(comadre_rep_flag,
                      check_NA_A == FALSE &
                      check_ergodic == TRUE)
```

**Question 1**

Show a histogram of the population growth rate, and a separate histogram of the generation time of your subset of species. Describe the patterns. Do they make sense to you?

```{r warning=FALSE}
#Calculating the population growth rate by finding the dominant eigenvalue of each matrix, then storing this in the reptiles dataframe
comadre_reptiles$lambda <- unlist(lapply(matA(comadre_reptiles), popbio::lambda))

#Calculating gen time using the rage function
comadre_reptiles$gen_T <- mapply(Rage::gen_time, matU(comadre_reptiles), matF(comadre_reptiles))

#ensuring a clean version of the dataset for phylogenies
comadre_reptiles_clean <- comadre_reptiles |>
  filter(!is.na(lambda), 
         !is.na(gen_T), 
         !is.infinite(gen_T))
```

```{r warning = F}
#Plotting the pop growth rate histogram - red abline represents a stable population that is neither growing nor shrinking
hist(comadre_reptiles_clean$lambda, xlab= "Population growth rate", main = NULL) 
  abline(v = 1, col = 'red', lty = 2, lwd = 2)
```

```{r warning = F}
#Plotting generation time histogram
hist(comadre_reptiles_clean$gen_T, xlab = "Generation time (years)", main = NULL)
```

The pop growth rate histogram shows that the majority of populations have a $\lambda$ close to 1, suggesting relatively stable population growth rates. There are, however, a few populations that have $\lambda$ values significantly lower than 1, suggesting a declining population, perhaps due to climate change. Conversely, there is a wide variety in populations with $\lambda$ values greater than one, with some populations growing steadily but others with explosive growth, such as the population with a $\lambda$ close to 2.5. As a result, the overall distribution is asymmetrical.

Similarly, most of the generation times are below 20 years, suggesting that most studied reptiles have short reproductive time-spans. The distribution is skewed, however, with some outlier populations that have longer (20-40 years) or significantly longer (50-80 years) generation times, but the frequency of such populations declines as the generation time increases.

**Question 2**

Fit a model to explore whether the generation time and population growth rate of your species’ populations are correlated. Would you have expected them to be correlated? Why?

```{r}

# Scatter plot
ggplot(comadre_reptiles_clean, aes(x = lambda, y = gen_T )) +
  geom_point(alpha = 0.7) +  
  labs(y = "Generation Time (gen_T)", 
       x = "Population Growth Rate (lambda)", 
       title = "Population Growth Rate vs. Scatter Plot of Generation Time") +
  theme_classic()

```

From the previous question and the above exploratory scatter plot, we can see that both $\lambda$ and generation time are **not** normally distributed. Therefore, we cannot use a linear model as this would violate the assumption of normality. Instead, we will use a spearmans correlation test:

```{r warning = F}
# Spearman's correlation test
spearman_result <- cor.test(comadre_reptiles_clean$lambda, 
                            comadre_reptiles_clean$gen_T, 
                            method = "spearman")

print(spearman_result)
```

Interestingly, we observe that there is no significant correlation as p \> 0.05, with a very weak negative correlation coefficient (rho) of -0.119. This is surprising as one might have expected species with higher population growth rates to have shorter generation times (r selected life strategy), which would hypothetically lead to a strong negative correlation. Given the lack of significance of the calculated correlation, however, this does not seem to be the case.

**Question 3**

Using the IUCN API token provided (or via your own), or via the IUCN csv file provided explore, using statistical models, whether generation time, on the one hand, and population growth rate, on the other hand, predict the Red List conservation status of your species subset.

```{r}
#Ensuring that the dataset only has one entry for each species, by picking the matrix based on the most recent publication
comadre_reptiles_clean_unique <- as.data.frame(comadre_reptiles_clean) |>  
  group_by(SpeciesAccepted) |>  
  slice_max(YearPublication, n = 1, with_ties = FALSE) |>  #Picking the matrix with highest publication year
  ungroup()

#Importing IUCN data
IUCN_data <- read.csv("IUCN_comadre_compadre.csv")

#Inserting the conservation status information back into the dataset
comadre_reptiles_iucn <- comadre_reptiles_clean_unique |>
  left_join(IUCN_data, by = "SpeciesAccepted") |>  # Merge datasets
  mutate(
    IUCNstatus = recode(IUCNstatus,  # Recode abbreviations to full names
      "EN" = "Endangered",
      "VU" = "Vulnerable",
      "NT" = "Near Threatened",
      "LC" = "Least Concern",
      .missing = "Not Assessed"  # Replace NA values
    ),
    IUCNstatus = factor(IUCNstatus, levels = c(
      "Endangered", "Vulnerable", "Near Threatened", "Least Concern", "Not Assessed"
    )),
    iucn_colour = recode(IUCNstatus,  # Assign colors directly
      "Endangered" = "Orange",
      "Vulnerable" = "Yellow",
      "Near Threatened" = "Dark green",
      "Least Concern" = "Green",
      "Not Assessed" = "Blue"
    ))

#One of the turtles, Podocnemis expansa, remains NA in the table as it has a unique conservation status - CD or conservation dependent. To rectify the NA, we will place it in the closest IUCN category.

comadre_reptiles_iucn$IUCNstatus[18] <- "Near Threatened"
comadre_reptiles_iucn$iucn_colour[18] <- "Dark green"
```

```{r}
ggplot(comadre_reptiles_iucn, aes(x = IUCNstatus, y = gen_T, 
                         fill = iucn_colour)) + 
  geom_boxplot(show.legend = F) +
  geom_jitter(width = 0.2, show.legend = F) +
  scale_fill_identity() +
  labs(x = "IUCN redlist status", y = "Generation time (years)") +
  theme_classic(base_size = 14)
```

```{r}
ANOVA_gen_T <- aov(gen_T~IUCNstatus, data = comadre_reptiles_iucn)
summary(ANOVA_gen_T)
```

***Generation Time:*** From the above box plot and ANOVA, we can see that generation time is not significantly predicted by conservation status (Pr\>F \>\> 0.05). The distribution is varied, with certain conservation categories such as least concern having individuals with both short and extremely long generation times.

```{r}
ggplot(comadre_reptiles_iucn, aes(x = IUCNstatus, y = lambda, 
                         fill = iucn_colour)) + 
  geom_boxplot(show.legend = F) +
  geom_jitter(width = 0.2, show.legend = F) +
  scale_fill_identity() +
  labs(x = "IUCN redlist status", y = "Population growth rate") +
  theme_classic(base_size = 14)
```

```{r}
ANOVA_gen_T <- aov(lambda~IUCNstatus, data = comadre_reptiles_iucn)
summary(ANOVA_gen_T)
```

***Growth Rate:*** Similarly, we can see that population growth rate is not significantly predicted by conservation status (Pr\>F \>\> 0.05). The distribution is more uniform, however, with most IUCN categories clustering around a $\lambda$ of one.

**Question 4**

Plot the values of generation time and of population growth rate of your species on the phylogeny provided to you (COMPADRE-COMADRE_Phylo_June_16_2019.tre). Is there any visual evidence of phylogenetic inertia in those traits? Why?

```{r}
#Load tree
tree <- read.tree("COMPADRE-COMADRE_Phylo_June_16_2019.tre")
#Clean tips
tree$tip.label <- gsub("_", " ", tree$tip.label)
#Prune tree to only keep tips that are in the reptiles dataframe
pruned_tree <- drop.tip(tree, setdiff(tree$tip.label, comadre_reptiles_iucn$SpeciesAccepted))

#Removing the species that are not represented as tips on the tree - "Sceloporus mucronatus mucronatus" & "Xenosaurus sp." and ordering the species

order_comadre_reptiles_iucn <- comadre_reptiles_iucn |>
  filter(SpeciesAccepted %in% pruned_tree$tip.label) |>   # Filter rows based on tip.label
  arrange(match(SpeciesAccepted, pruned_tree$tip.label)) # Reorder the rows to match tip.label order

row.names(order_comadre_reptiles_iucn) <- order_comadre_reptiles_iucn$SpeciesAccepted

```

```{r}
#Calculating log lambdas for the tips to be used in phylogeny
log_lambdas <- log(setNames(order_comadre_reptiles_iucn$lambda, rownames(order_comadre_reptiles_iucn)))

# Using a contMap to reconstruct ancestral traits for population growth rate
reptile_contMap_lambda <- contMap(pruned_tree, log_lambdas,plot=FALSE,res=200)
reptile_contMap_lambda <- setMap(reptile_contMap_lambda, c("white","#FFFFB2","#FECC5C","#FD8D3C","#E31A1C"))
plot(reptile_contMap_lambda, fsize=c(0.7,0.8), leg.txt="log(population growth rate)")
par(mar=c(5.1,4.1,4.1,2.1)) ## reset margins to default
```

```{r}
#Repeating the process for generation time
log_gen_T <- log(setNames(order_comadre_reptiles_iucn$gen_T, rownames(order_comadre_reptiles_iucn)))
reptiles_contMap_gen_T <- contMap(pruned_tree, log_gen_T,plot=FALSE,res=200)
reptiles_contMap_gen_T <- setMap(reptiles_contMap_gen_T, c("white","light blue","blue","violet","purple"))
plot(reptiles_contMap_gen_T, fsize=c(0.7,0.8), leg.txt="log(generation time) (years)")
```

***Population growth rate:*** The phylogeny shows that most ancestral nodes are predicted to have relatively even growth rates around 1, with faster growing species evolving very suddenly towards the present. This is paradoxical - suggesting high phylogenetic inertia in the past followed by significantly lower inertia today, but can be explained by sampling bias - the ancestors with very low growth rates were predisposed to go extinct, so it is likely growth rate has always been a trait of low inertia, but we lack the sampling to show this.

***Generation time:*** The phylogenetic tree shows high levels of inertia, with clades such as crocodylus maintaining long generation times throughout history. Clades that have evolved shorter generation times, on the other hand, such as Xenosaurus or Sceloporus, are unlikely to recover longer generation times as changing a vital life history trait like lifespan is very difficult.

**Question 5**
