---
title: "Big Data Computer Assessment"
output:
  html_document: default
  pdf_document: default
---

Candidate Number: 1073601

```{r, echo=FALSE}
# List of required packages
required_packages <- c(
  "mosaic", "phytools", "popbio", "popdemo", "Rage", "Rcompadre",
  "remotes", "rotl", "taxize", "tidyverse")

# Installs the missing packages
missing_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(missing_packages)) {
  install.packages(missing_packages)
}
#Loads the missing packages
lapply(required_packages, library, character.only = TRUE)
```

```{r}
#Downloading the Comadre dataset
#Fetch the most recent version
comadre <- cdb_fetch("comadre")

#Selecting Reptiles
#located worldwide and not from captivity:
comadre_rep <- subset(comadre,
                   # Consider only reptiles
                   Class == "Reptilia" & 
                   # anywhere in the world so no lat requirement
                   # studied under "control" (i.e. no treatment) conditions
                   MatrixTreatment == "Unmanipulated" &
                   # and only from "W"ild populations
                   MatrixCaptivity == "W") 
#Flag and remove matrices with missing values or dont meet assumptions of ergodicity
comadre_rep_flag <- cdb_flag(comadre_rep)
comadre_reptiles <- subset(comadre_rep_flag,
                      check_NA_A == FALSE &
                      check_ergodic == TRUE)
```

**Question 1**

Show a histogram of the population growth rate, and a separate histogram of the generation time of your subset of species. Describe the patterns. Do they make sense to you?

```{r warning=FALSE}
#Calculating the population growth rate by finding the dominant eigenvalue of each matrix, then storing this in the reptiles dataframe
comadre_reptiles$lambda <- unlist(lapply(matA(comadre_reptiles), popbio::lambda))

#Calculating gen time using the rage function
comadre_reptiles$gen_T <- mapply(Rage::gen_time, matU(comadre_reptiles), matF(comadre_reptiles))

#ensuring a clean version of the dataset for phylogenies
comadre_reptiles_clean <- comadre_reptiles |>
  filter(!is.na(lambda), 
         !is.na(gen_T), 
         !is.infinite(gen_T))
```

```{r warnings = F}
#Plotting the pop growth rate histogram - red abline represents a stable population that is neither growing nor shrinking
hist(comadre_reptiles_clean$lambda, xlab= "Population growth rate", main = NULL) 
  abline(v = 1, col = 'red', lty = 2, lwd = 2)
```

```{r warnings = F}
#Plotting generation time histogram
hist(comadre_reptiles_clean$gen_T, xlab = "Generation time (years)", main = NULL)
```

The pop growth rate histogram shows that the majority of populations have a $\lambda$ close to 1, suggesting relatively stable population growth rates. There are, however, a few populations that have $\lambda$ values significantly lower than 1, suggesting a declining population, perhaps due to climate change. Conversely, there is a wide variety in populations with $\lambda$ values greater than one, with some populations growing steadily but others with explosive growth, such as the population with a $\lambda$ close to 2.5.

Similarly, most of the generation times are below 20 years, suggesting that most studied reptiles have short reproductive time-spans. The distribution is skewed, however, with some outlier populations that have longer (20-40 years) or significantly longer (50-80 years) generation times, but the frequency of such populations declines as the generation time increases.

**Question 2**

Fit a model to explore whether the generation time and population growth rate of your
species’ populations are correlated. Would you have expected them to be correlated? Why?

